
### .hand[Your turn!]

- Clone the repository from the course webpage
- Read in the `nobels.csv`.
- Split into two (STEM and non-STEM):
  - Create a new data frame, `nobel_stem`, that filters for the STEM fields
(Physics, Medicine, Chemistry, and Economics).
  - Create another data frame, `nobel_nonstem`, that filters for the remaining
fields.
- Write out the two data frames to `nobel-stem.csv` and `nobel-nonstem.csv`,
respectively, to `data/`.


### .hand[Your turn!]
- open `sales-excel.Rmd`. 
- Load the `sales.xlsx` file from the `data-raw/` folder, using appropriate arguments for the `read_excel()` function such that it looks like the output on the left.
- **Stretch goal:** Manipulate the sales data such that it looks like the output on the right.

.pull-left[
```{r echo=FALSE}
sales = read_excel("l03/data/sales.xlsx", skip = 3, col_names = c("id", "n"))
sales
```
]
.pull-right[
```{r echo=FALSE}
sales %>%
  mutate(
    is_brand_name = str_detect(id, "Brand"),
    brand = if_else(is_brand_name, id, NA_character_)
  ) %>%
  fill(brand) %>%
  filter(!is_brand_name) %>%
  select(brand, id, n) %>%
  mutate(
    id = as.numeric(id),
    n = as.numeric(n)
  )
```
]

---
