---
title: "BST430 Lecture 02-C"
subtitle: "Memory and Environments in R"
author: "Seong-Hwan Jun"
institute: "U of Rochester"
date: today
format: 
  revealjs: 
    theme: simple
    slide-number: true
---

## Namespace

- Sometimes there are name clashes. For example, `filter()` exists in both `dplyr` and `stats`.
- In such cases, we can specify the package name using `::` operator.

## Namespace

```{r, echo=TRUE}
stats::filter
```

## Namespace

```{r, echo=TRUE}
dplyr::filter
```

## Global environment

- When you start R, the global environment is created automatically.
- You can see the objects in your global environment using `ls()`.
- You can remove objects from your global environment using `rm()`.
- The global environment is the top-level environment in R, and it has no parent environment.

## Local environment 

- Most typically created when a function is called.
- Variables declared within a local environment is only accessible within that environment.
- When a function is called, R creates a new local environment for that function call.
- When the function exits, the local environment is destroyed.
- Local environments can access variables in the global environment (parent environment), but not vice versa.

## Names and values

```{r, echo=TRUE}
x <- c(1, 2, 3)
```

This line allocates a space in memory for the vector `c(1, 2, 3)` and assigns the name `x` to that space.

`x` is a reference to an actual object, in this case, vector `c(1, 2, 3)`.

## Names and values

```{r, echo=TRUE}
y <- x
```

This will create another name, `y`, that also binds to the existing vector `c(1, 2, 3)`.

## Names and values

Print their memory addresses:

```{r, echo=TRUE}
print(lobstr::obj_addr(x))
print(lobstr::obj_addr(y))
```

They point to the same object.

## Names and values

By the way, functions also have names and values.

```{r, echo=TRUE}
print(lobstr::obj_addr(mean))
```

Functions are also objects in R.

- `mean` is a name that points to a function object.
- The name `mean` is just a reference to that object. The actual function object or definition is stored in memory.

## Copy-on-modify

R employs a copy-on-modify strategy to manage memory efficiently.

When you modify an object, R creates a copy of the object only if it is necessary (i.e., if there are multiple references to the same object).

## Copy-on-modify

We will use `tracemem` function, which tracks when an object is copied in memory.

```{r, echo=TRUE}
xxx <- c(1, 2, 3)
tracemem(xxx)
yyy <- xxx # Copy does not occur here.
print(lobstr::obj_addr(xxx))
print(lobstr::obj_addr(yyy))
```

## Copy-on-modify

```{r, echo=TRUE}
yyy[1] <- 100 # copy occurs here!
print(xxx)
print(yyy)
print(lobstr::obj_addr(xxx))
print(lobstr::obj_addr(yyy))
```

## What happens if you modify `xxx`?

```{r, echo=TRUE}
xxx[1] <- -100
print(xxx)
print(lobstr::obj_addr(xxx))
```

:::{.incremental}
- What happened to the original object `c(1, 2, 3)`?

- It is still in memory, but now there are no names pointing to it, so it will be garbage collected if you run out of space.
:::

## Copy on function calls?

```{r, echo=TRUE}
f <- function(a) {
  lobstr::obj_addr(a)
  return(a)
}
xyz <- c(1, 10, 100)
lobstr::obj_addr(xyz)
ret_obj <- f(xyz)
lobstr::obj_addr(ret_obj)
```

No copy takes place because `xyz` is not modified inside function `f`.

## Where does copy take place?

```{r, echo=TRUE}
f <- function(a) {
  lobstr::obj_addr(a)
  a[1] <- -1
  return(a)
}
xyz <- c(1, 10, 100)
lobstr::obj_addr(xyz)
ret_obj <- f(xyz)
lobstr::obj_addr(ret_obj)
```

## Modify-in-place

If only a single reference exists for an object, then modifying is to occur in place (no copy is made) -- but this is nearly impossible to check.

- When we call `tracemem(x)`, R seems to create a new reference to the object `x`, so now there are two references.
- Even when references are removed, the way R keeps track of references is complex, so it is hard to be sure how many references there are.

An exception: environments. These are a special type of object and always modified in place.

## Performance implications

Why is frequent copying harmful for performance?

:::{.incremental}
- Copying large objects takes time and memory.
- Frequent copying can lead to increased memory usage and slower performance.
:::

## Shallow vs Deep copy

- Shallow copy: creates a new object that references the same memory as the original object.
- Deep copy: creates a new object with its own copy of the data in memory.

## Lists: `list`

```{r, echo=TRUE}
l1 <- list(1, 2, 3)
```

- Each element of the list is a reference to an object.
- e.g., `l1[[1]]` is a reference to object `1` stored in memory and `l1[[2]]` is a reference to object `2` and so on.

## Lists: `list`

```{r, echo=TRUE}
l2 <- l1 # Shallow copy here.
lobstr::ref(l1)

l2[[3]] <- 4
lobstr::ref(l2)
```

:::{.incremental}
- `l2[[1]]` and `l2[[2]]` still point to the same object as `l1[[1]]` and `l1[[2]]`.
- `l2[[3]]` now points to a new object `4` in memory.
:::

## Data frames: `data.frame`

- Data frame is a list of vectors -- each column points to a vector stored in memory.
- What happens if you modify a column? a row? -> Find out in the lab.

## Character vectors

```{r, echo=TRUE}
x <- c("hi", "hello", "a", "b", "hi")
```

The first and the last element should point to distinct objects both containing the string "hi"... right?

## Character vectors

```{r, echo=TRUE}
x <- c("hi", "hello", "a", "b", "hi")
lobstr::ref(x, character = TRUE)
```

R uses a global pool of strings to avoid storing multiple copies of the same string in memory.

