---
title: "BST430 Lecture 06-B"
subtitle: "Reference based data manipulation with data.table"
author: "Seong-Hwan Jun"
format: revealjs
---

## R objects are immutable

- When we modify a `data.frame` (or `tibble`), a copy is made.
- This is inefficient for large datasets.
- We may want to pass around the data structure to functions where it modifies in place without making a copy.

## Enter `data.table`

- `data.table` is an R package that provides an enhanced version of `data.frame`.
- I started using it because I wanted to do *reference-based* data manipulation.

What is reference-based data manipulation?

## Modify in place via `:=`

`:=` is a special operator in `data.table` that allows you to modify data by reference.

```{r echo=TRUE}
library(data.table)
DT = data.table(x = 1:5, y = letters[1:5])
tracemem(DT)
DT[,y:= toupper(y)]
DT
```

No memory copy.

## What about `data.frame`?

```{r echo=TRUE}
DF = data.frame(x = 1:5, y = letters[1:5])
tracemem(DF)
DF$y <- toupper(DF$y)
```

## Passing by reference

Pass data tables to functions and modify them in place without making a copy.

```{r echo=TRUE}
set.seed(123)
num_elements <- 10000
DT = data.table(x = sample(letters, num_elements, replace = TRUE), y = sample(letters, num_elements, replace = TRUE))
DT
```

## Passing by reference

```{r echo=TRUE}
modify_DT = function(DT) {
  DT[x == 'i' & y == 'j', y := "ij"]
}
tracemem(DT)
modify_DT(DT) # Does not return anything.
DT[x=='i' & y == 'ij']
```

## Passing by reference

```{r echo=TRUE}
lobstr::obj_addr(DT)
```

## Modifying a subset of rows

```{r echo=TRUE}
DF = data.frame(x = sample(letters, num_elements, replace = TRUE), y = sample(letters, num_elements, replace = TRUE))
tracemem(DF)
DF[DF$x=='i' & DF$y == 'i','y'] <- "ij"
lobstr::obj_addr(DF)
```

## Delete a column

```{r echo=TRUE}
DT[, y := NULL]
DT
```

## Input and output

```{r echo=TRUE}
DT <- fread("l05/data/trump/approval_polllist.csv")
temp <- DT[,.(avg_approval=mean(approve)),by=.(population)]
fwrite(temp, "l05/data/trump/approval_by_population.csv")
```

Their input and output functions are pretty versatile and just "works" in most cases.

## Primary keys

- `data.table` supports primary keys for fast subsetting.
- Set a key using `setkey()`.

