---
title: "BST430 Lecture 06-B"
subtitle: "Reference based data manipulation with data.table"
author: "Seong-Hwan Jun"
format: revealjs
---

## R objects are immutable

- When we modify a `data.frame` (or `tibble`), a copy is made.
- This is inefficient for large datasets.
- We may want to pass around the data structure to functions where it modifies in place without making a copy.

## `data.frame` example

```{r echo=TRUE}
DF = data.frame(x = 1:5, y = letters[1:5])
DF
tracemem(DF) # This will print memory address of DF when it is copied.
DF$y <- toupper(DF$y)
```

## `data.table` and `:=` operator

An R package that provides an enhanced version of `data.frame`; it allows modify in place.

```{r echo=TRUE}
library(data.table)
DT = data.table(x = 1:5, y = letters[1:5])
tracemem(DT)
DT[,y:= toupper(y)]
DT
```

No memory copy.

## Passing by reference

`data.table` can be passed to functions that perform modification in place.

```{r echo=TRUE}
set.seed(123)
num_elements <- 10000
DT = data.table(x = sample(letters, num_elements, replace = TRUE), 
                y = sample(letters, num_elements, replace = TRUE))
DT
```

## Passing by reference

```{r echo=TRUE}
modify_DT = function(DT) {
  DT[x == 'i' & y == 'j', y := "ij"]
}
tracemem(DT)
modify_DT(DT) # Does not return anything.
DT[x=='i' & y == 'ij']
```

## Passing by reference

Check that the memory address of DT remains the same after modification.

```{r echo=TRUE}
lobstr::obj_addr(DT)
```

## Modifying a subset of rows

```{r echo=TRUE}
DF = data.frame(x = sample(letters, num_elements, replace = TRUE), y = sample(letters, num_elements, replace = TRUE))
tracemem(DF)
DF[DF$x=='i' & DF$y == 'i','y'] <- "ij"
lobstr::obj_addr(DF)
```

## Delete a column

```{r echo=TRUE}
DT[, y := NULL]
DT
```

## Input and output

```{r echo=TRUE}
DT <- fread("l05/data/trump/approval_polllist.csv")
temp <- DT[,.(avg_approval=mean(approve)),by=.(population)]
fwrite(temp, "l05/data/trump/approval_by_population.csv")
```

Their input and output functions are pretty versatile and just "works" in most cases.

## Primary keys

- `data.table` supports primary keys for fast subsetting.
- Set a key using `setkey()`.
- This allows fast subsetting via binary search `O(log n)` instead of `O(n)`.

## Notes

Fully supports features needed for data manipulation:

- table joins.
- reshaping (longer and wider).
- can be used with `dplyr` and `ggplot2` -- much of what we learned for `tibble` applies to `data.table`

`tidyverse` yields more readable code. `data.table` syntax is a bit more terse and cryptic.

## References

- `data.table` vignettes: <https://rdatatable.gitlab.io/data.table/index.html>
