---
title: "Lab6 Key"
author: "Andrew McDavid"
date: "10/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-packages-data, message = FALSE, eval = TRUE}
library(tidyverse)
library(lubridate)
library(ggrepel)
library(vroom)
```

## Ex1

```{r}
cases = read_csv('data/us-states.csv')
```

```{r}
sub_counties = filter(cases, state %in% c("Alabama", "Arizona", 'California', 'Florida', 'New York', 'Texas'), date >= as.Date('2020-03-01'))
sub_county_long = pivot_longer(sub_counties, c(cases, deaths), names_to = 'stat')
plt = ggplot(sub_county_long, aes(x = date, y = value, color = stat)) + 
  geom_line(alpha = .5, position = 'stack') + facet_wrap(~state+stat, scales = 'free_y') + theme_minimal() +
  scale_color_manual(values = c('purple', 'brown')) + theme(axis.text.x = element_text(angle = 45))
plt
```


## Ex2

```{r}
library(lubridate)
plt + xlim(as.Date(c('2021-08-01', '2021-10-01')))
```

There's strong ~7 day periodicity in the data, likely relating to reporting delays over the weekend.  Not all states feature this pattern, however.

## Ex3

```{r}
sub_counties = sub_counties %>% mutate(day_of_week = weekdays(date))
weekdays_plot = ggplot(filter(sub_counties, date > as.Date('2021-07-01')), aes(x = fct_reorder(weekdays(date), cases), y = cases))+ geom_boxplot() + facet_grid(state ~ month(date), scales = 'free') + theme(axis.text.x = element_text(angle = 90))
weekdays_plot
```

Plotting the number of cases by state, by day-of-week, by month shows that not every state has this feature so strongly.  Florida is especially bad -- in July it reported no data on Saturday-Monday.  Oddly, this changed in August.  Texas reports no data on Sundays. Arizona, New york and Alabama show some day-of-week variability but don't have data dropping out entirely.

## Ex4

```{r}
sub_county_long = pivot_longer(sub_counties, c(cases_avg_per_100k, deaths_avg_per_100k), names_to = 'stat')
plt %+% sub_county_long #+ geom_vline(data = annot, aes(xintercept = date))# = quantity), size = 2, color = 'black')
```

## Ex5

1.  `group_by(date)`, sum, then slice_max.
2.  `group_by(date)`, var, then slice_max.
3.  `pivot_wider`, then a filter operation.

## Ex6 and 7

```{r}
rowVar = function(matrix, na.rm = FALSE){
  apply(matrix, 1, var, na.rm = na.rm)
}
```

```{r}
case_mat = reshape2::acast(sub_counties, date ~ state, value.var = "cases_avg_per_100k", fill = 0)
which.max(rowSums(case_mat))
which.max(rowVar(case_mat))
which(case_mat[,'California'] > case_mat[,'Texas'] & case_mat[,'California'] < case_mat[, 'Florida'])
```


## Ex 8

```{r}
cor_state = cor(case_mat)
pheatmap::pheatmap(cor_state)
```




```{r, eval = FALSE}
pc_stat = sub_counties %>% group_by(date) %>% summarize(total_pc = sum(cases_avg_per_100k), var_pc = var(cases_avg_per_100k))

max_total = slice_max(pc_stat, total_pc) %>% mutate(quantity = 'Highest Cases p.c.')
max_var =  slice_max(pc_stat, var_pc) %>% mutate(quantity = 'Highest Variance p.c.')

sub_county_long = pivot_longer(sub_counties, c(cases_avg_per_100k, deaths_avg_per_100k), names_to = 'stat')
annot = bind_rows(max_total, max_var) #%>% left_join(sub_county_long)
plt %+% sub_county_long + geom_vline(data = annot, aes(xintercept = date))# = quantity), size = 2, color = 'black')
```

