---
title: "Lab 06 - Covid time series - one"
output: 
  tufte::tufte_html:
    css: ../lab.css
    tufte_variant: "envisioned"
    highlight: pygments
link-citations: yes
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE)
```


# Learning goals

-  Exploratory data analysis
-  Manipulating matrices

# Getting started

Go to the course GitHub organization and locate your repo, clone it in RStudio and open the R Markdown document.
Knit the document to make sure it compiles without errors.

## Warm up

Before we introduce the data, let's warm up with some simple exercises.
Update the YAML of your R Markdown file with your information, knit, commit, and push your changes.
Make sure to commit with a meaningful commit message.
Then, go to your repo on GitHub and confirm that your changes are visible in your Rmd **and** md files.
If anything is missing, commit and push again.

## Packages

We'll use the **tidyverse**, as well as (likely) **lubridate**, **pheatmap**, and possibly **reshape2**.
These packages are already installed for you.
You can load them by running the following in your Console:

```{r eval = TRUE, message = FALSE}
library(reshape2)
library(tidyverse)
library(lubridate)
```

```{r}
smooth_n = function(x, n = 7) stats::filter(x, rep(1/n, n), method = 'convolution', sides = 1)

smoothed = sub_counties %>%
  group_by(state) %>% mutate(across(c(cases_new, deaths_new), smooth_n))

waves =  as.Date(c("2020-03-01", "2020-06-01", "2020-09-15", "2021-04-01", "2021-07-01", "2021-11-01"))

plt %+% (smoothed %>% pivot_longer(c(cases_new, deaths_new), names_to = 'stat')) + geom_vline(data = tibble(date = waves), aes(xintercept = date))

```





## Part II

ACF and waves


```{r}
cases_new = sub_counties %>% pivot_wider(date, names_from = state, values_from = cases_new)
cases_mat = cases_new %>% select(-date) %>% as.matrix()
rownames(cases_mat) = as.character(cases_new$date)
```

*  Upon what day did the greater total number of cases per capita occur?
*  Upon what day was the intra-state variance the greatest?  Add these with geom_text_repel by setting a new data argument.
*  Upon what days did California have more cases p.c. than Texas, but less than Florida?
*  Convert cases and deaths to a wide matrix, repeat previous questions (don't regenerate plots)
*  Generate the correlation matrix of cases and deaths and visualize using a heatmap.







3.  A commonly-stated rule of thumb is that deaths lag cases by 14 days.  Construct a matrix of smoothed cases vs smooth deaths for the following periods: 
*  Wave 1 California
*  Wave 1 New York
*  Wave 1 Florida
*  Wave 2 New York
*  Wave 2 Florida
*  Wave 3 New York
*  Wave 3 Arizona
*  Wave 3 Alabama
*  Wave 5 New York
*  Wave 5 Alabama
*  Wave 5 California


```{r}
library(broom)
states = smoothed %>% group_by(state, date) %>% summarize(across(c(cases_new, deaths_new), sum)) %>% ungroup()
states = states %>% mutate(wave = cut(date, breaks =waves))

plt %+% (states %>% pivot_longer(c(cases_new, deaths_new), names_to = 'stat')) + aes(group = NULL)
acfs = states %>% group_by(wave, state) %>% summarize(acf = list(ccf(cases_new, deaths_new, na.action = na.pass, lag.max = 60, plot = FALSE)))

acfs2 = acfs %>% rowwise() %>% mutate(td = list(tidy(acf))) %>% select(-acf) %>% unnest(cols = c(td))

ggplot(acfs2, aes(x = lag, y = acf)) + facet_grid(wave ~ state) + geom_line()


acfs_max = acfs2 %>% group_by(state, wave) %>% summarize(max_lag = lag[which.max(acf)])
```


```{r}
states_j = left_join(acfs_max, states)

(states_j %>% pivot_longer(c(cases_new, deaths_new), names_to = 'stat')) %>%
  ggplot(aes(x = date, y = value, group = stat, color = max_lag)) + geom_line() + 
  facet_grid(state ~ wave, scales = 'free') + scale_color_viridis_c()
```





`r emo::ji("broom")` 🧶 ✅ ⬆️ Lint, Knit, *commit, and push your changes to GitHub with an appropriate commit message. Make sure to commit and push all changed files so that your Git pane is cleared up afterwards and review the md document on GitHub to make sure you're happy with the final state of your work.*

# Wrapping up

Go back through your write up to make sure you're following coding style guidelines we discussed in class.
Make any edits as needed.

Also, make sure all of your R chunks are properly labelled, and your figures are reasonably sized.

Once the team leader for the week pushes their final changes, others should *pull* the changes and knit the R Markdown document to confirm that they can reproduce the report.
