---
title: "L16  Multivariate Regression - AE"
author: "Andrew McDavid"
date: "11/7/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(broom)
library(GGally)
data(mtcars)
```

## Part i (fit and infer)

0. Using GGally, make a pairs plot of `disp`, `wt`, and `am`.

1. Using the `mtcars` data, fit a linear model of displacement and `am` (as a factor) on weight. 

2. What are the regression coefficient estimates? Interpret them.

3. Now, fit a linear model of displacement, `am` (as a factor), **and the interaction** between the two variables on weight.  What are the regression coefficient estimates?  Interpret them.

4. Using `glance`, evaluate evidence in favor of the interaction in (3) vs the additive model in (2) using a metric of your choice.

5. (If you have time) make a plot showing the regression coefficient estimates from (3), and their 95\% confidence intervals.

## Part ii (prediction)

1.  From the BAC dataset, suppose someone had 5 beers. Use
  `predict()` to predict their BAC.
  
```{r, eval = FALSE, echo = FALSE}
library(openintro)
data("bac")
```

## Part iii (many linear models)
  
1.  Using `gapminder`, fit a linear model on `log(pop)` vs years since 1990 over all countries.  Interpet the slope and intercept in terms of the unlogged variables.

```{r}
library(gapminder)
```


2.  Now, completing the code below, fit this model per country.  Report the countries with the fastest and slowest population growth rates.

```{r, error = TRUE}
fit_pop = function(){
  ##Fill in this function
}

many_fits = gapminder %>% 
  group_by(country) %>%
  summarize(fit = fit_pop(across()))  %>%
  ungroup()

filter(many_fits, country == 'India')$fit[[1]]
filter(many_fits, country == 'Italy')$fit[[1]]
```


```{r, error = TRUE}
many_fits %>% 
  rowwise() %>%
  mutate(tidy_out = list(tidy(___))) %>%
  unnest(___)
```

