---
title: "BST430  Lecture 5b"
subtitle: "dplyr ii"
author: "Andrew McDavid"
institute: "U of Rochester"
date: "2021-09-12 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css:
      - default
      - css/lexis.css
      - css/lexis-fonts.css
    lib_dir: libs
    seal: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      titleSlideClass: ['ur-title', 'center', 'middle']
      ratio: "3:2"
---
class: middle

```{r child = "setup.Rmd"}
```
```{r, include = FALSE}
library(tidyverse)
```

.hand[# A lil' bit more dplyr]

---

## Agenda

* grouping by multiple variables
* Empty groups and summarizing
* ungrouping
* grouped mutates and filters
* special selection operators
* repeating operations on multiple columns with `across`
<!-- * row names and row numbers-->

---
class: code70

## grouping

As we saw in lecture 3 by adding `group_by()` to a pipeline, we stratify by a categorical variable when we `summarize()`:
.pull-left[
**Grand margin**
```{r}
library(nycflights13)
ontime = flights %>% 
  mutate(ontime = arr_delay <= 0)  
ontime %>% 
  summarize(ontime_pct = mean(ontime, na.rm = TRUE)*100)
```
]

.pull-right[
**stratified by carrier**
```{r}
ontime %>% group_by(carrier) %>% 
  summarize(ontime_pct = mean(ontime, na.rm = TRUE)*100) #<<
```
]

---

## `group_by()` multiple variables

It's easy to group by multiple variables:

```r
flights %>% group_by(<var_1>, <var_2>, ..., <var_n>) %>%
   summarize()
```

---

## `group_by()` caveat

Be aware that this only returns the combinations of var_1, var_2, ... that exist in the data. This can lead to erroneous calculations if you are joining tables downstream. To get the Cartesian product:

1.  Convert `var1`, `var2`, ... to factors with `factor`.<sup>1</sup>
2.  Set  `group_by(..., .drop = FALSE)`.

.footnote[[1] We'll discuss the pain and pleasure<sup>2</sup> of factors in greater detail next week.

[2] If you are a masochist.]

---

## Nonsense from dropping

.panelset[

.panel[.panel-name[Code]
```{r hourly-departures-drop, fig.show='hide'}
departures = flights %>% group_by(carrier, hour) %>% summarize(n_departures = n())
ggplot(departures, aes(x = hour, y = n_departures)) + 
  geom_line() + 
  facet_wrap(~carrier, scales = 'free_y') #<<
```
]

.panel[.panel-name[Plot]
```{r ref.label="hourly-departures-drop", echo = FALSE, out.width="75%"}
```
]
]

---

## Keep (wacky factors)

.panelset[

.panel[.panel-name[Code]
```{r hourly-departures-keep, fig.show='hide'}
departures2 = flights %>% 
  mutate(carrier = factor(carrier), hour = factor(hour)) %>% #<<
  group_by(carrier, hour, .drop = FALSE) %>% #<<
  summarize(n_departures = n())
plt = ggplot(departures2, aes(x = hour, y = n_departures, group = carrier)) + 
  geom_line() + 
  facet_wrap(~carrier, scales = 'free_y')
plt
```
]

.panel[.panel-name[Plot]
```{r ref.label="hourly-departures-keep", echo = FALSE, out.width="75%"}
```
]
]

---

## Keep (preserve original)

.panelset[

.panel[.panel-name[Code]
```{r hourly-departures-keep-hour, fig.show='hide', message = TRUE}
departures3 = flights %>% 
  mutate(carrier = factor(carrier), hourf = factor(hour)) %>%
  group_by(carrier, hourf, .drop = FALSE) %>%
  summarize(n_departures = n()) %>%
  mutate(hour = as.numeric(as.character(hourf))) #<<
(plt %+% departures3) + aes(x = hour) #<<
```
]

.panel[.panel-name[Plot]
```{r ref.label="hourly-departures-keep-hour", echo = FALSE, out.width="75%"}
```
]
]

---

## grouping is contagious

Note the message:

>  `summarise()` has grouped output by 'carrier'. You can override using the `.groups` argument.

This is telling us that by default, the output will be grouped if we use it in another pipeline.

To save screen real estate, I'm going to turn off this message going forward:
```{r}
options(dplyr.summarise.inform = FALSE)
```

In general you should set `.groups` explicitly.

---

## `ungroup()` to remove grouping variables.

.pull-left[
Default
```{r}
ontime_drop = ontime %>% group_by(carrier, dest) %>% summarize(ontime_pct = mean(ontime, na.rm = TRUE)*100)
ontime_drop %>% summarize(n_dest = n_distinct(dest))
```
]

.pull-right[
ungrouped
```{r}
ontime_drop %>% ungroup() %>% #<<
  summarize(n_dest = n_distinct(dest))
```
]
---

## `group_by() filter()`

We can combine group_by with filter, mutate or arrange to evaluate these verbes on each group.

```{r}
ontime_1500 = ontime %>% group_by(dest) %>% filter(n() > 1500) %>%
  mutate(n_flights_dest = n(), .after = 1)

ontime_1500 %>% arrange(n_flights_dest)

```

---

## Carrier performance by destination

```{r, fig.width = 10, fig.asp=.5}
ontime_drop = ontime_1500 %>% group_by(dest, carrier) %>%  
  summarize(ontime_pct = mean(ontime, na.rm  = TRUE)*100)
ggplot(ontime_drop, aes(y = ontime_pct, x = dest)) + geom_point() + coord_flip() + theme_minimal(base_size = 8)
```

.question[What's with the carriers with 0% or 100% ontime percentages for a destination?]

---

## The "law" of small numbers

```{r}
ontime_drop = ontime_1500 %>% group_by(dest, carrier) %>%  
  summarize(ontime_pct = mean(ontime, na.rm  = TRUE)*100,
            n_departed = sum(!is.na(ontime)))

ggplot(ontime_drop, aes(y = ontime_pct, x = dest, color = n_departed<10)) + geom_point() + coord_flip() + theme_minimal(base_size = 8)
```

.question[What are some statistical fixes to this issue?]
---

## Without small numbers

```{r}
ontime_drop_10 = filter(ontime_drop, n_departed >= 10)
ggplot(ontime_drop_10, aes(y = ontime_pct, x = dest)) + geom_boxplot() + 
  coord_flip()  + theme_minimal(base_size = 8)
```

Better...but it's hard to see the rank order of routes.

---

## A sensible plot order

```{r}
 ontime_drop_10 = ontime_drop_10 %>% ungroup() %>% 
  mutate(dest = forcats::fct_reorder(dest, ontime_pct)) #<<
ggplot(ontime_drop_10, aes(x = dest, y = ontime_pct)) + geom_boxplot() + 
  coord_flip()  + theme_minimal(base_size = 8)
```

We'll discuss the `fct_reorder` business in greater detail next week.

---

## `select()` 201

`select()`, `pivot_wider()`, `pivot_longer()`, and friends use [tidy-select](https://dplyr.tidyverse.org/reference/dplyr_tidy_select.html), which is a meta-language to select columns from a data frame. There are several helper functions you can use within `select()`:

-   `starts_with("abc")`: matches names that begin with "abc",
-   `ends_with("xyz")`: matches names that end with "xyz",
-   `contains("ijk")`: matches names that contain "ijk",
-   `num_range("x", 1:3)`: matches `x1`, `x2` and `x3`,
-   `last_col(offset = 0)`: last column, minus offset,
-   `where(<function>)`: columns where `<function>` returns `TRUE`,
-   `:` range, `c` concatenate, `!` negate, `&` intersect
-   ...and a few others.
---

## select delay columns

```{r}
flights %>% select(ends_with('delay'))
```
---

## everything except columns containing "time"

```{r}
flights %>% select(!contains('time'))
```
---

## character columns

```{r}
flights %>% select(where(is.character))
```

---

## Columns with no missing values

```{r}
flights %>% select(where(~!any(is.na(.x))))
```
The above uses a tidyverse way to define an **anonymous function** and is equivalent to 
```
none_na = function(x) !any(is.na(x))
```

---
class: middle

.hand[# Repeating an operation with `across`]

---

## `across()`

If you need to repeat the same operation on multiple columns, you can avoid many keystrokes by using across.

It's syntax is a bit idiomatic, but in general, it looks something like
```r
verb(across(.cols = <tidy selection>, 
            .funs = list(<fun1>, <fun2>, ...)))
```
where `verb` is `summarize`,  `mutate`, and a few others where unquoted columns are provided to a transformation.

---

## NYC AirBNB data

```{r, include = FALSE}
nycbnb = read_csv("https://urmc-bst.github.io/bst430-fall2021-site/hw_lab_instruction/hw-01-airbnb/data/nylistings.csv")
```

```{r}
nycbnb
```

---

## Nights

.panelset[

.panel[.panel-name[Code]
```{r nights-plot, fig.show='hide'}
availability = nycbnb %>% group_by(neighborhood) %>% 
  summarize(across(c(maximum_nights, minimum_nights, availability_90), median)) #<<
ggplot(availability, 
       aes(x = neighborhood, 
           ymin = minimum_nights,  
           y = availability_90, 
           ymax = maximum_nights)) + 
  geom_pointrange() + coord_flip() + theme_minimal() +  
  labs(x = NULL, y = 'Availability (minimum, 90 day, maximum)')
```
]

.panel[.panel-name[Plot]
```{r, ref.label='nights-plot',  echo = FALSE, out.width="75%"}
```
]
]
---

## Transform by variable type

This can be very powerful:
```{r}
median_mad = nycbnb %>% group_by(neighborhood) %>% 
  summarise(across(where(is.numeric), #<<
                   list(median = median, mad = mad))) #<<
glimpse(median_mad)
```
